Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-06-25 overview.tex
Many small changes to all files -- GB final edits

The comoving total fluid energy density $E$ is given by
\begin{equation}
E =  e + \frac{\rho v^2}{2}  + \frac{B^2}{2a}
        \label{eq:total_energy_def},
\end{equation}
where $e$ is the comoving thermal energy density. The total comoving
isotropic pressure $p^*$ is given by
\begin{equation}
p^* = p + \frac{B^2}{2a},
\end{equation}
and the quantity $p$ is the thermal pressure.  We use units such that the magnetic
permeability is unity ($\mu_0=1$).  The equations are closed by an
equation of state and Poisson's equation (in the Newtonian limit) for
the gravitational potential $\phi$:
%
\begin{eqnarray}
%
  e &=& \frac{p}{(\gamma - 1)},
  \label{eq:eq_of_state} \\
%
  \nabla^2 \phi &=& 4 \pi G \left( \rho_{\rm total} - \rho_0 \right)
%                   + 3 p_{total}/c^2) - \Lambda. 
  \label{eq:potential}
\end{eqnarray}
%
The equation of state is shown here for an ideal gas with a ratio of
specific heats $\gamma$.  The gravitational potential $\phi$ is
sourced by the total mass density contrast, where $\rho_{\rm total} =
\rho_{\rm gas} + \rho_{\rm dm} + \rho_{\rm stars}$ and $\rho_0$ is the
mean density.  Although we write the equations including both magnetic
field terms and comoving coordinates, the code is frequently used both
in the purely hydrodynamic limit ($\vecB = 0$, referred to as HD
below) and without cosmological expansion ($a = 1$, $\dot{a} = 0$).

@@

\begin{eqnarray}
\vecx & = & \vecx' / a, \\
\vecv & = & a \hspace{0.5mm} \dot{\vecx} = 
              \vecvp - \dot{a}\vecx, \\
\rho  & = & a^3 \rhop,   \\
p     & = & a^3 \pp, \\
E     & = & a^3 \left(\Ep - 
            \dot{a} \, \vecx \cdot \vecvp - 
            \frac{1}{2} \dot{a}^2 \vecx^2\right), \\
\phi  & = & \phip + \frac{1}{2} \, a \ddot{a} \, \vecx^2, \\
\vecB & = & a^2 \vecB^\prime,
\end{eqnarray}
where primes indicate quantities in proper coordinates (fixed frame).
We note that the definition of comoving $\vecB$ used here is such
that a uniform field is constant in a homogeneous expanding universe
and is also the quantity used in the MHD-CT solver described
below; however, it is not universal,
and in particular is slightly different from that used both in \citet{Li08a}
and used in the MHD-Dedner solver (see section~\ref{sec.num.hydro-muscl}
for more details).
%These definitions result in the following comoving equivalents to eqs. 
%(\ref{eq:mass})-(\ref{eq:potential}).
%%
%
%\begin{eqnarray} 
%\frac{\partial \rhop}{\partial t} 
%          + \frac{1}{a} \vecvp \cdot \grad \rhop 
%    & = & - \frac{1}{a} \rhop \div \vecvp, 
%        \label{eq:mass_prime} \\
%%
%\frac{\partial \vecvp}{\partial t}  
%          + \frac{1}{a} ( \vecvp \cdot \grad ) \vecvp 
%    & = & - \frac{\dot{a}}{a} \vecvp 
%          - \frac{1}{a \rhop} \grad \pp 
%          - \frac{1}{a} \grad \phip, 
%        \label{eq:momentum_prime} \\
%%
%\frac{\partial \Ep}{\partial t} 
%         + \frac{1}{a} \vecvp \cdot \grad \Ep 
%   & = & - \frac{\dot{a}}{a} (3 \frac{\pp}{\rhop} + {\vecvp \cdot \vecvp} )
%         - \frac{1}{a \rhop } \div (p \vecvp) 
%%\nonumber   & &   %uncomment for 2-column
%         - \frac{1}{a} \vecvp \cdot \grad \phip 
%                + \Gamma - \Lambda.
%       \label{eq:total_energy_prime} \\
%%
%\Ep  & = & \ep + \frac{1}{2} {\vecvp \cdot \vecvp},
%        \label{eq:total_energy_def_prime} \\
%%
%\ep                  & = & 
%        \pp/\left[ \left( \gamma - 1 \right) \rhop \right],
%      \label{eq:state_prime} \\ 
%{\nabla}^2 \phip    & = & 
%        \frac{4\pi G}{a} ( \rhop_{total} - \rhop_0 ).
%      \label{eq:poisson}
%\end{eqnarray}
%%
%%We use the proper peculiar baryonic velocity $\vecv_b \equiv a(t) d\vecx / dt$, proper pressure $p$, and modified gravitational potential $ \phi $ which is related to the potential in proper coordinates $\Phi$ by 
%%$$\phi \equiv \Phi + \frac{1}{2} a \ddot{a} \vecx^2. $$
%%The density, however, is comoving:
%%$$\rho_b \equiv \rho_{b,proper}a(t)^3, $$
The expansion parameter $a \equiv 1/(1 + z)$ follows the expansion of
a smooth, homogeneous background, where $z$, the redshift, is a
function only of $t$.  All spatial derivatives are determined with
respect to the comoving position $\vecx$, which removes the universal
expansion from the coordinate system. The evolution of $a(t)$ is
governed by the second Friedmann equation for the expansion of a
spatially homogeneous and isotropic universe
%
\begin{eqnarray}
\frac{\ddot{a}}{a} & = & 
      - \frac{4 \pi G }{3 a^3 } (\rho_0 
      + 3p_0/c^2) 
      + \Lambda_c c^2 / 3 .
      \label{eq:expansion} 
\end{eqnarray}
%
Here $\rho_0$ is the mean comoving mass density (including both
baryonic and dark matter), $p_0$ is the comoving background pressure
contribution, and $\Lambda_c$ is the cosmological constant.
%$\gamma$ is the ratio of specific heats, and $\Lambda$ is the cosmological constant.
%We note that if $\gamma$ is not $5/3$ then $E$ in the first term on
%the right side of eq.~(ref{eq:total_energy}) must be replaced with 
%eq.~(\ref{eq:state}).
This system of equations is limited to the non-relativistic regime and
assumes that curvature effects are not important --- both assumptions
are reasonable as long as the size of the simulated region is small
compared to the radius of curvature and the Hubble length $c/H$, where
$c$ is the speed of light and $H = \dot{a}/a$ is the Hubble constant.

@@

In addition, \enzo\ can solve the mass conservation equations for a set of
chemical species and their reactions.  For any species $i$ with comoving
number density $n_i$, these equations have the form:
\begin{equation}
  \frac{\partial n_i}{\partial t} 
  + \frac{1}{a} \div (n_i \vecv) = 
  \sum_j k_{ij}(T) \, n_i n_j + \sum_j \Gamma^{\rm ph}_j n_j 
  \label{eq:species_evolution}
\end{equation}
where $k_{ij}$ are the rate coefficients for the two-body reactions
and are usually functions of only the gas species (we will
specifically note the cases where we either include three-body
reactions or have density-dependent rates).  The $\Gamma^{\rm ph}_j$
are destruction/creation rates due to photoionizations and/or
photodissociations.  Currently the species \enzo\ can follow include
H, H$^+$, He, He$^+$, He$^{++}$, and e$^-$, and through additional
options also H$^-$, H$_2$, H$_2^+$, and HD, D, and D$^+$. Lastly,
\enzo\ can also track the advection of one or more comoving metal density fields,
which can contribute to the radiative cooling and star formation
processes.

The code can include either a homogeneous radiation background 
or evolve an inhomogeneous
radiation field either by directly solving the radiative transfer
equation along rays or by solving a set of moment equations derived
from the radiative transfer equation. In comoving coordinates
\citep[e.g.,][]{Gnedin97} the radiative transfer equation reads
%
\begin{equation}
  \label{eq:rteqn}
  \frac{1}{c} \; \frac{\partial I_\nu}{\partial t} + 
  \frac{a_{\rm em}}{a} \, \hat{n} \cdot \grad I_\nu -
  \frac{H}{c} \; \left( \nu \frac{\partial I_\nu}{\partial \nu} -
  3 I_\nu \right) = -\kappa_\nu I_\nu + j_\nu .
\end{equation}
%
Here $I_\nu \equiv I(\nu, \mathbf{x}, \Omega, t)$ is specific
intensity of the radiation, with dimensions of energy per time $t$ per
solid angle $\Omega$ per unit area per frequency $\nu$.  The second
term represents the propagation of radiation, where the factor $a_{\rm
  em}/a$ accounts for cosmic expansion since the time of emission.
The third term describes both the cosmological redshift and dilution
of radiation.  On the right hand side, the first term captures
absorption ($\kappa_\nu \equiv \kappa_\nu(\mathbf{x},\nu,t)$), and the
second term emission ($j_\nu \equiv j_\nu(\mathbf{x},\nu,t)$ from any
sources of radiation (whether point or diffuse).

@@

Finally, \enzo\ implements the equations of isotropic heat conduction
in a manner similar to that of \citet{2007ApJ...664..135P}, where the isotropic heat flux is given by
%
\begin{equation}
\fcond = -\kappa_{\rm cond} \grad T.
\end{equation}
%
Here $\kappa_{\rm cond} = f_{\rm sp} \, \kappa_{\rm sp}$ is the heat
conduction coefficient, given as a fraction $f_{\rm sp}$ of the
Spitzer conductivity $\kappa_{\rm sp}$ \citep{1962pfig.book.....S},
and T is the gas temperature (with fluids explicitly assumed to be
single-temperature).  Saturation of the heat flux in high temperature,
low density regimes (such as the intracluster medium in galaxy
clusters) is taken into account.

Thermal conduction in a plasma can be strongly affected by the
presence of magnetic field lines, which may strongly suppress heat
flow perpendicular to the magnetic field.  To include this physics, it
is possible to allow heat transport only along (not across) magnetic
field lines, as follows:

\begin{equation}
\fcond= -\kappa_{\rm cond} \,\myvec{b} ( \myvec{b} \cdot \grad T),
\end{equation}
where $\myvec{b}$ is the unit vector denoting the direction of
the magnetic field.

@@

Given the hierarchy at some time $t$, we advance the solution in the
manner of a W-cycle in a multigrid solver.  First, we determine the
maximum timestep allowed for the coarsest grid based on a variety of
accuracy and stability criteria and advance the grid by that time
interval, $\Delta t_0$.  We then move down to the next level and
advance all the grids on that level by a timestep $\Delta t_1 \leq
\Delta t_0$, which is the minimum of all the allowed timesteps for
those grids.  If there are more levels, we repeat this procedure until
the bottom level of the hierarchy has been reached.  Once there, we
continue advancing the grids on the lowest level until they have
``caught up'' to the coarser parent grids (i.e., $\sum \Delta t_l =
\Delta t_{l-1}$).  This procedure repeats itself until all grids have
been advanced by a total time of $\Delta t_0$ (see
Figure~\ref{fig:wcycle} for a schematic).

\begin{figure}
\begin{center}
%\includegraphics[width=0.6\textwidth]{figures/wcycle.eps}
\includegraphics[width=0.6\textwidth]{figures/timestepping.eps}
\end{center}
\caption{\emph{Left:} Example of the timesteps on a 2-level AMR
  hierarchy.  \enzo\ does not restrict the timesteps on the finer levels
  to be a factor of $1/2^n$ smaller.  \emph{Right:} The order in which
  the AMR grids are evaluated on each level.\vspace{1ex}}
\label{fig:wcycle}
\end{figure}

@@

As described in \citet{1995CoPhC..89..149B}, we modify the method for use in
hypersonic flows when the thermal energy $e$ is extremely small
compared to the total energy $E$.  This situation presents a problem
because in the total energy method the temperature is computed by
subtracting one large number from another (i.e. the kinetic energy
from the total energy), which tends to generate large numerical
inaccuracies. We overcome this difficulty by additionally solving a
thermal energy equation and using $e$ from this equation when we
expect the error to be large.

@@

The current implementation of self-gravity in \enzo\ uses a Fast
Fourier technique \citep{Hockney88} to solve Poisson's equation on the
root grid on each timestep.  The advantage of using this method is
that it is fast, accurate, and naturally allows both periodic and
isolated boundary conditions for the gravity, choices which are very
common in astrophysics and cosmology.  On subgrids, we interpolate the
boundary conditions from the parent grid (either the root grid or some
other subgrid). The Poisson equation is then solved on every timestep
using a multigrid technique on one subgrid at a time. Aside from
self-consistently calculating the gravitational potential arising from
the baryon fields and particles in the simulation, there are also a
number of options for specifying static gravitational fields, and in
fact including self-gravity is optional.

@@

Although the particles are fixed in mass once initialized (with the
exception of star particles, which can lose mass in the feedback
process), we can create them with any initial set of masses and
positions.  For example, in many cosmological simulations static
subgrids are included from the beginning in order to improve the
initial spatial and baryonic mass resolution, and these subgrid are
populated with lower-mass particles to correspondingly improve the
collisionless mass resolution.
%% One particle per initial grid point seems to provide approximately
%% equal sampling between the dark matter and gas. \mqk{This
%% contradicts the conclusions of O'Shea et al. 2005 that an initial
%% grid twice as fine as the mean particle separation is necessary to
%% properly capture the abundance of low mass halos.}

>>>>

The comoving total fluid energy density $E$ is given by
\begin{equation}
E =  e + \frac{\rho v^2}{2}  + \frac{B^2}{2a}
        \label{eq:total_energy_def},
\end{equation}
where $e$ is the comoving thermal energy density. The total comoving
isotropic pressure $p^*$ is given by
\begin{equation}
p^* = p + \frac{B^2}{2a},
\end{equation}
and the quantity $p$ is the thermal pressure.  We use units such that the magnetic
permeability is unity ($\mu_0=1$).  The equations are closed by an
equation of state and Poisson's equation for the gravitational potential $\phi$:
%
\begin{eqnarray}
%
  e &=& \frac{p}{(\gamma - 1)},
  \label{eq:eq_of_state} \\
%
  \nabla^2 \phi &=& \frac{4 \pi G}{a} \left( \rho_{\rm total} - \rho_0 \right)
%                   + 3 p_{total}/c^2) - \Lambda. 
  \label{eq:potential}
\end{eqnarray}
%
The equation of state is shown here for an ideal gas with a ratio of
specific heats $\gamma$.  The gravitational potential $\phi$ is
sourced by the total mass density contrast, where $\rho_{\rm total} =
\rho_{\rm gas} + \rho_{\rm dm} + \rho_{\rm stars}$ and $\rho_0$ is the
mean density.  Although we write the equations including both magnetic
field terms and comoving coordinates, the code is frequently used both
in the purely hydrodynamic limit ($\vecB = 0$, referred to as HD
below) and without cosmological expansion ($a = 1$, $\dot{a} = 0$).

@@

\begin{eqnarray}
\vecx & = & \vecx' / a, \\
\vecv & = & a \hspace{0.5mm} \dot{\vecx} = 
              \vecvp - \dot{a}\vecx, \\
\rho  & = & a^3 \rhop,   \\
p     & = & a^3 \pp, \\
E     & = & a^3 \left(\Ep - 
            \dot{a} \, \vecx \cdot \vecvp - 
            \frac{1}{2} \dot{a}^2 \vecx^2\right), \\
\phi  & = & \phip + \frac{1}{2} \, a \ddot{a} \, \vecx^2, \\
\vecB & = & a^2 \vecB^\prime,
\end{eqnarray}
where primes indicate quantities in proper coordinates (fixed frame).
We note that the definition of comoving $\vecB$ used here is such
that a uniform field is constant in a homogeneous expanding universe
and is also the quantity used in the MHD-CT solver described
below; however, it is not universal,
and is slightly different from that used both in \citet{Li08a}
and in the MHD-Dedner solver (see section~\ref{sec.num.hydro-muscl}
for more details).
%These definitions result in the following comoving equivalents to eqs. 
%(\ref{eq:mass})-(\ref{eq:potential}).
%%
%
%\begin{eqnarray} 
%\frac{\partial \rhop}{\partial t} 
%          + \frac{1}{a} \vecvp \cdot \grad \rhop 
%    & = & - \frac{1}{a} \rhop \div \vecvp, 
%        \label{eq:mass_prime} \\
%%
%\frac{\partial \vecvp}{\partial t}  
%          + \frac{1}{a} ( \vecvp \cdot \grad ) \vecvp 
%    & = & - \frac{\dot{a}}{a} \vecvp 
%          - \frac{1}{a \rhop} \grad \pp 
%          - \frac{1}{a} \grad \phip, 
%        \label{eq:momentum_prime} \\
%%
%\frac{\partial \Ep}{\partial t} 
%         + \frac{1}{a} \vecvp \cdot \grad \Ep 
%   & = & - \frac{\dot{a}}{a} (3 \frac{\pp}{\rhop} + {\vecvp \cdot \vecvp} )
%         - \frac{1}{a \rhop } \div (p \vecvp) 
%%\nonumber   & &   %uncomment for 2-column
%         - \frac{1}{a} \vecvp \cdot \grad \phip 
%                + \Gamma - \Lambda.
%       \label{eq:total_energy_prime} \\
%%
%\Ep  & = & \ep + \frac{1}{2} {\vecvp \cdot \vecvp},
%        \label{eq:total_energy_def_prime} \\
%%
%\ep                  & = & 
%        \pp/\left[ \left( \gamma - 1 \right) \rhop \right],
%      \label{eq:state_prime} \\ 
%{\nabla}^2 \phip    & = & 
%        \frac{4\pi G}{a} ( \rhop_{total} - \rhop_0 ).
%      \label{eq:poisson}
%\end{eqnarray}
%%
%%We use the proper peculiar baryonic velocity $\vecv_b \equiv a(t) d\vecx / dt$, proper pressure $p$, and modified gravitational potential $ \phi $ which is related to the potential in proper coordinates $\Phi$ by 
%%$$\phi \equiv \Phi + \frac{1}{2} a \ddot{a} \vecx^2. $$
%%The density, however, is comoving:
%%$$\rho_b \equiv \rho_{b,proper}a(t)^3, $$
The expansion parameter $a \equiv 1/(1 + z)$ follows the expansion of
a smooth, homogeneous background, where $z$, the redshift, is a
function only of $t$.  All spatial derivatives are determined with
respect to the comoving position $\vecx$, which removes the universal
expansion from the coordinate system. The evolution of $a(t)$ is
governed by the second Friedmann equation for the expansion of a
spatially homogeneous and isotropic universe
%
\begin{eqnarray}
\frac{\ddot{a}}{a} & = & 
      - \frac{4 \pi G }{3 a^3 } (\rho_0 
      + 3p_0/c^2) 
      + \Lambda_c c^2 / 3 .
      \label{eq:expansion} 
\end{eqnarray}
%
Here $\rho_0$ is the mean comoving mass density (including both
baryonic and dark matter), $p_0$ is the comoving background pressure
contribution, and $\Lambda_c$ is the cosmological constant.
%$\gamma$ is the ratio of specific heats, and $\Lambda$ is the cosmological constant.
%We note that if $\gamma$ is not $5/3$ then $E$ in the first term on
%the right side of eq.~(ref{eq:total_energy}) must be replaced with 
%eq.~(\ref{eq:state}).
This system of equations is limited to the non-relativistic regime and
assumes that curvature effects are not important --- both assumptions
are reasonable as long as the size of the simulated region is small
compared to the radius of curvature and the Hubble length $c/H$, where
$c$ is the speed of light and $H = \dot{a}/a$ is the Hubble constant.

@@

In addition, \enzo\ can solve the mass conservation equations for a set of
chemical species and their reactions.  For any species $i$ with comoving
number density $n_i$, these equations have the form:
\begin{equation}
  \frac{\partial n_i}{\partial t} 
  + \frac{1}{a} \div (n_i \vecv) = 
  \sum_j k_{ij}(T) \, n_i n_j + \sum_j \Gamma^{\rm ph}_j n_j 
  \label{eq:species_evolution}
\end{equation}
where $k_{ij}$ are the rate coefficients for the two-body reactions
and are usually functions of only temperature (we will
specifically note the cases where we either include three-body
reactions or have density-dependent rates).  The $\Gamma^{\rm ph}_j$
are destruction/creation rates due to photoionizations and/or
photodissociations.  Currently the species \enzo\ can follow include
H, H$^+$, He, He$^+$, He$^{++}$, and e$^-$, and through additional
options also H$^-$, H$_2$, H$_2^+$, and HD, D, and D$^+$. Lastly,
\enzo\ can also track the advection of one or more comoving metal density fields,
which can contribute to the radiative cooling and star formation
processes.

The code can include either a homogeneous radiation background 
or evolve an inhomogeneous
radiation field either by directly solving the radiative transfer
equation along rays or by solving a set of moment equations derived
from the radiative transfer equation. In comoving coordinates
\citep[e.g.,][]{Gnedin97} the radiative transfer equation reads
%
\begin{equation}
  \label{eq:rteqn}
  \frac{1}{c} \; \frac{\partial I_\nu}{\partial t} + 
  \frac{a_{\rm em}}{a} \, \hat{n} \cdot \grad I_\nu -
  \frac{H}{c} \; \left( \nu \frac{\partial I_\nu}{\partial \nu} -
  3 I_\nu \right) = -\kappa_\nu I_\nu + j_\nu .
\end{equation}
%
Here $I_\nu \equiv I(\nu, \mathbf{x}, \Omega, t)$ is specific
intensity of the radiation, with dimensions of energy per time $t$ per
solid angle $\Omega$ per unit area per frequency $\nu$.  The second
term represents the propagation of radiation, where the factor $a_{\rm
  em}/a$ accounts for cosmic expansion since the time of emission.
The third term describes both the cosmological redshift and dilution
of radiation.  On the right hand side, the first term captures
absorption ($\kappa_\nu \equiv \kappa_\nu(\mathbf{x},\nu,t)$), and the
second term emission ($j_\nu \equiv j_\nu(\mathbf{x},\nu,t)$) from any
sources of radiation (whether point or diffuse).

@@

Finally, \enzo\ implements the equations of isotropic heat conduction
in a manner similar to that of \citet{2007ApJ...664..135P}, where the isotropic heat flux is given by
%
\begin{equation}
\fcond = -\kappa_{\rm cond} \grad T.
\label{eq:conduction}
\end{equation}
%
Here $\kappa_{\rm cond} = f_{\rm sp} \, \kappa_{\rm sp}$ is the heat
conduction coefficient, given as a fraction $f_{\rm sp}$ of the
Spitzer conductivity $\kappa_{\rm sp}$ \citep{1962pfig.book.....S},
and T is the gas temperature (with fluids explicitly assumed to be
single-temperature).  Saturation of the heat flux in high temperature,
low density regimes (such as the intracluster medium in galaxy
clusters) is taken into account.

Thermal conduction in a plasma can be strongly affected by the
presence of magnetic field lines, which may strongly suppress heat
flow perpendicular to the magnetic field.  To include this effect, it
is possible to allow heat transport only along (not across) magnetic
field lines, as follows:

\begin{equation}
\fcond= -\kappa_{\rm cond} \,\myvec{b} ( \myvec{b} \cdot \grad T),
\label{eq:anisotropic_conduction}
\end{equation}
where $\myvec{b}$ is the unit vector denoting the direction of
the magnetic field.

@@

Given the hierarchy at some time $t$, we advance the solution in the
manner of a W-cycle in a multigrid solver.  First, we determine the
maximum timestep allowed for the coarsest grid based on a variety of
accuracy and stability criteria and advance the grid by that time
interval.  We then move down to the next level and
advance all the grids on that level.  If there are more levels, we repeat this procedure until
the bottom level of the hierarchy has been reached.  Once there, we
continue advancing the grids on the lowest level until they have
``caught up'' to the coarser parent grids.  This procedure repeats itself until all grids have
been advanced by the desired time.

@@

As described in \citet{1995CoPhC..89..149B}, we modify the method for use in
hypersonic flows when the thermal energy $e$ is extremely small
compared to the total energy $E$.  This situation presents a problem
because in the total energy method the thermal energy is computed by
subtracting one large number from another (i.e. the kinetic energy
from the total energy), which tends to generate large numerical
inaccuracies. We overcome this difficulty by additionally solving a
thermal energy equation and using $e$ from this equation when we
expect the error to be large.

@@

The current implementation of self-gravity in \enzo\ uses a Fast
Fourier technique \citep{Hockney88} to solve Poisson's equation on the
root grid on each timestep.  The advantage of using this method is
that it is fast, accurate, and naturally allows both periodic and
isolated boundary conditions for the gravity, choices which are very
common in astrophysics and cosmology.  On subgrids, we interpolate the
boundary conditions from the parent grid (either the root grid or some
other subgrid). The Poisson equation is then solved on every timestep
using a multigrid technique on one subgrid at a time. Aside from
self-consistently calculating the gravitational potential arising from
the baryon fields and particles in the simulation, there are also a
number of options for specifying static gravitational fields.

@@

Although the particles are fixed in mass once initialized (with the
exception of star particles, which can lose mass in the feedback
process), we can create them with any initial set of masses and
positions.  For example, in many cosmological simulations static
subgrids are included from the beginning in order to improve the
initial spatial and baryonic mass resolution, and these subgrids are
populated with lower-mass particles to correspondingly improve the
collisionless mass resolution.
%% One particle per initial grid point seems to provide approximately
%% equal sampling between the dark matter and gas. \mqk{This
%% contradicts the conclusions of O'Shea et al. 2005 that an initial
%% grid twice as fine as the mean particle separation is necessary to
%% properly capture the abundance of low mass halos.}

### 2013-06-25 parallel.tex
Many small changes to all files -- GB final edits

Broadly, excellent scaling in \enzo's unigrid mode is seen for a
variety of problems as long as each compute core is given an adequate
amount of work to do.  For cosmological simulations, this value has
been empirically determined to be roughly $128^3$ cells per core.  If
fewer cells per core are used, the CPU is essentially data-starved 
and poor scaling is observed due to computing units being idle while
waiting for information to be communicated from other processes (for,
e.g., boundary information or gravity solves).  Substantially larger
cell counts per core would in principle help scaling by reducing the
amount of inter-process communication needed, but larger cell counts
are typically impractical on most machines due to the amount of memory
available on each node.

As a final point, we observe that scaling at larger core counts has
been measured, but only with an experimental hybrid-parallel (MPI +
OpenMP) version of \enzo.  Using this version, scaling comparable to
that shown in Figure~\ref{fig.uniscale} was seen on up to 98,304 cores
on the NICS Kraken XT5 (an upgraded version of the XT4 machine used
for the scaling study shown in the figure), using 2-8 OpenMP threads
per MPI process.  Hybrid parallelism has the potential to
substantially improve scaling by reducing the amount of communication
per grid tile, as described in the previous paragraph; however, the
optimal ratio of OpenMP threads per MPI task seems to vary
substantially between computational platforms and astrophysical
problems/required \enzo\ physics modules, and thus we hesitate to
provide strict guidelines here.

@@

In this section, we develop a simple, approximate model to estimate
the computational time and memory required to complete a given
calculation.  Note that we are {\it not} trying to model parallel
performance (which was discussed in the previous section), but have an
even simpler goal: to determine how to scale computational time
estimates as we increase the simulation resolution.  This is a
straightforward thing to do in codes that use static grids, as the
computational effort per timestep is constant.  However, for an AMR
calculation, the number of grids that will be generated during the run
is not known in advance, and so the CPU time per problem time can vary
drastically throughout the calculation.  For example, at the beginning
of the simulation when the densities are nearly uniform, only the
static grid is required and the calculation progresses rapidly.
However, as structure forms and dense clumps are generated, the number
of grid points swells by orders of magnitude (an increase of $10^3$ is
not uncommon) and most of the CPU time is consumed at late physical
time in the simulation.
Therefore, simply performing a few steps at the beginning of the
calculation does not produce a good estimate of the required CPU time.

@@

As, the figure demonstrates, the $f_l$ values all show a sharp decline
with level, dropping as power-laws with the level $l$.  Focusing first
on the ``refine-everywhere" cosmological runs, which both show similar
behavior despite the different box sizes and redshifts at which we
collect the data.  In fact, we find these results are very typical for
cosmological runs, with only slight variations depending on the exact
problem that is being simulated.  The top-left panel
shows the ratio of $f_{l}/f_{l+1}$ which, remarkably, hovers around
the critical value of 0.0625 determined earlier.  This implies that
the relative CPU usage of each level, shown in the lower-left panel,
is nearly flat, and that going to additional levels of refinement only
increases the amount of calculation by a factor of roughly $1/l_{\rm
max}$.  Even in the most extreme case, the L80 run, the last level
adds only 25\% to the computation time.

@@

These conclusions are, however, highly dependent on the type of run
being performed.  We contrast this cosmological case to the simulation
of a galaxy formation run, as shown by the dot-dashed curves in
Figure~\ref{fig:scaling}.  Note that grid levels 1 and 2 in this simulation are statically
refined to ensure that the Lagrangian volume of the galaxy halo is resolved
to at least level 2 at all times. More importantly, the
$f_{l+1}/f_{l}$ ratio of the highest levels are systematically
slightly larger than the critical 0.0625 value, indicating that the
CPU time is dominated by the most refined levels, as shown in the
bottom-left panel.  The memory usage is dominated by levels 2 and 3,
the static levels, as the bottom-right panel demonstrates.  In this
case, adding more levels of refinement would somewhat increase the
cost of the simulation.

We have focused thus far on scaling when increasing $l_{\rm max}$, but
without changing the refinement criteria.  However, we can also keep
the levels fixed and modify the refinement criteria.  For cosmological
runs, this is typically done by increasing the mass resolution (and
simultaneously decreasing the dark matter particle mass), and adding
additional linear small-scale modes to the initial conditions.  The
effect of this is to boost the $f_l$ values at all levels (except for
the root grid where $f_0$ is already 1) by a factor linearly
proportional to the increase in the mass resolution.  This is because
the refinement criteria typically boosts the number of refined cells
on the first and subsequent levels by this factor.  Again, we have
empirically demonstrated that this scaling is approximately valid,
provided that we keep the maximum level constant.

>>>>

Broadly, excellent scaling in \enzo's unigrid mode is seen for a
variety of problems as long as each compute core is given an adequate
amount of work to do.  For cosmological simulations, this value has
been empirically determined to be roughly $128^3$ cells per core.  If
fewer cells per core are used, the CPU is essentially data-starved 
and poor scaling is observed due to computing units being idle while
waiting for information to be communicated from other processes (for,
e.g., boundary information or gravity solves).  Substantially larger
cell counts per core would in principle help scaling by reducing the
amount of inter-process communication needed.

As a final point, we observe that scaling at larger core counts has
been measured, but only with an experimental hybrid-parallel (MPI +
OpenMP) version of \enzo.  Using this version, scaling comparable to
that shown in Figure~\ref{fig.uniscale} was seen on up to 98,304 cores
on the NICS Kraken XT5 (an upgraded version of the XT4 machine used
for the scaling study shown in the figure), using 2-8 OpenMP threads
per MPI process.  Hybrid parallelism has the potential to
substantially improve scaling by reducing the amount of communication
per grid tile, as described in the previous paragraph.

@@

In this section, we develop a simple, approximate model to estimate
the computational time and memory required to complete a given
calculation.  Note that we are {\it not} trying to model parallel
performance (which was discussed in the previous section), but have an
even simpler goal: to determine how to scale computational time
estimates as we increase the simulation resolution.  This is a
straightforward thing to do in codes that use static grids, as the
computational effort per timestep is constant.  However, for an AMR
calculation, the number of grids that will be generated during the run
is not known in advance, and so the CPU time per problem time can vary
drastically throughout the calculation.  For example, at the beginning
of a cosmological simulation, when the densities are nearly uniform, only the
static grid is required and the calculation progresses rapidly.
However, as structure forms and dense clumps are generated, the number
of grid points swells by orders of magnitude (an increase of $10^3$ is
not uncommon) and most of the CPU time is consumed at late physical
time in the simulation.
Therefore, simply performing a few steps at the beginning of the
calculation does not produce a good estimate of the required CPU time.

@@

As, the figure demonstrates, the $f_l$ values all show a sharp decline
with level, dropping as power-laws with the level $l$.  Focusing first
on the ``refine-everywhere" cosmological runs, which both show similar
behavior despite the different box sizes and redshifts at which we
collect the data.  In fact, we find these results are very typical for
cosmological runs, with only slight variations depending on the exact
problem that is being simulated.  The top-right panel
shows the ratio of $f_{l}/f_{l+1}$ which, remarkably, hovers around
the critical value of 0.0625 determined earlier.  This implies that
the relative CPU usage of each level, shown in the lower-left panel,
is nearly flat, and that going to additional levels of refinement only
increases the amount of calculation by a factor of roughly $1/l_{\rm
max}$.  Even in the most extreme case, the L80 run, the last level
adds only 25\% to the computation time.

@@

These conclusions are, however, highly dependent on the type of run
being performed.  We contrast this cosmological case to the simulation
of a galaxy formation run, as shown by the dot-dashed curves in
Figure~\ref{fig:scaling}.  Note that grid levels 1 and 2 in this simulation are statically
refined to ensure that the Lagrangian volume of the galaxy halo is resolved
to at least level 2 at all times. More importantly, the
$f_{l+1}/f_{l}$ ratio of the highest levels are systematically
slightly larger than the critical 0.0625 value, indicating that the
CPU time is dominated by the most refined levels, as shown in the
bottom-left panel.  The memory usage is dominated by levels 2 and 3,
the static levels, as the bottom-right panel demonstrates.  In this
case, adding more levels of refinement would substantially increase the
cost of the simulation.

We have focused thus far on scaling when increasing $l_{\rm max}$, but
without changing the refinement criteria.  However, we can also keep
the levels fixed and modify the refinement criteria.  For cosmological
runs, this is typically done by increasing the mass resolution (and
simultaneously decreasing the dark matter particle mass), and adding
additional linear small-scale modes to the initial conditions.  The
effect of this is to boost the $f_l$ values at all levels (except for
the root grid where $f_0$ is already 1) by a factor linearly
proportional to the increase in the mass resolution.  This is because
the refinement criteria typically boosts the number of refined cells
on the first and subsequent levels by this factor.  Again, we have
empirically tested that this scaling is approximately valid,
provided that we keep the maximum level constant.
